Synthesizing Quotient Lenses

### 2018-03-10 papers/QuotientLenses/main.tex
partial pass over background/example section; all buyt section on QRE Lenses subsection

\subsection{Quotient Lenses}
Unfortunately, while {\em Optician} greatly simplifies the task of
programming bijective string lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Lovelace, Ada'' vs ``Ada Lovelace.''

When two instances of the same format differ only terms of such inessential
information, it is desirable
to treat them as identical and to map them into canonical
representations.
The alternative is to require extra machinery to preserve the
inessential details, which leads to complex and brittle lenses.
Continuing our example, the following two \bibtex{} citations represent the same
logical object and as a result, should be translated to the same EndNote record.
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title},
}

@Book{     Lovelace,
Title = {Generic
Title},
Author = "Lovelace, Ada",     }
\end{verbatim}

One natural way to define lenses that will synchronize data in these formats
is to provide the following items:
\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
  \bibtex{} and EndNote. Such descriptions may be given by regular
  expressions, for instance.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats.
  \item A bijective string lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}. Of course, some of these elements are optional.  For
instance, it is not necessary to document internal and external formats.
However, we do view it as good engineering practice to do so, as such
documentation makes lenses easier to read, understand, use and maintain.

The first way in which our work simplifies Boomerang's quotient lenses is
that we merge the first three pieces of quotient lenses that we just described
into a single concept, {\em Quotient Regular Expression}, or QREs. A single QRE
enables a programmer to add annotations to the outer regular expression; these
annotations can then be used to infer an inner regular expression, as well as a
canonizer from the outer regular expression to the inner one. In the following
subsection, we describe a set of QRE combinators that we have integrated into
Boomerang and give an example of how a programmer might define a QRE
specification of \bibtex{} using these combinators.

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
To construct a QRE specification of a \bibtex{} format, we might first define a
whitespace format, which, externally, matches any non-zero number of whitespace
characters, and converts any such whitespace into a single space character (its
canonical form). This whitespace-normalizing QRE is easily defined by using the
\kw{project} primitive: \bcp{Why is it called ``project''?}

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Lovelace, Ada''
and that the \codefont{space\_name} format describes ``Ada Lovelace''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{c\_to\_s} canonization function used by squash
is simply to write it from scratch in some ordinary programming language.
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{c\_to\_s} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
let c_to_s = get l
\end{lstlisting}

The first line above synthesizes a lens between \codefont{comma\_name} and
\codefont{space\_name} using the listed example transformation as a guide.  The
second line simply extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

Another QRE primitive is the sequential composition combinator ``;''.  For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \kw{squash} QRE and a QRE that projects away
the extra whitespace in between \codefont{first\_name} and
\codefont{last\_name}.

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp+) $\to$ ",\n")
\end{lstlisting}

The final QRE combinator is the normalize combinator, which is also implemented
in Boomerang. This combinator allows a programmer to manually define a function
$f$ which sends each string that matches a regular expression $R$ to some
canonical representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence relation
defined by the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that
match $R$, $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p\_wsp} which matches any non-zero number of whitespace
characters and converts any such whitespace into a single space character can
be expressed using the normalize combinator as
\lstinline{normalize (wsp+, " ", f)}.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective string lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

>>>>

{\em Optician} greatly simplifies the task of programming bijective
string lenses, but not all bidirectional transformations are
bijective.  For instance, \bibtex{} users are not typically interested
in preserving whitespace between words.  The order of author and title
fields is also likely irrelevant, and there may be equivalent ways of
writing the same name: ``Lovelace, Ada'' vs ``Ada Lovelace.''
Consequently, the following two \bibtex{} citations represent the same
logical object even though they differ in nonessential details.
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Notes},
}

@Book{     Lovelace,
Title = {Notes},
Author = "Lovelace, Ada",     }
\end{verbatim}

When mapping these records into another format, such as EndNote, we
must decide what to do with the nonessential information.  A bijective
mapping must preserve all the information, including the extraneous
details, which leads to complex and brittle lenses complicated by the
need to preserve unimportant details.  
A better approach is to identify records that differ only in the
nonessential information, mapping them into a canonical representation.
This canonical representation is then mapped into the target format.
With this approach, both of the above \bibtex{} records would be
mapped to the same EndNote record.

We use {\em Quotient Regular Expresions} (QREs) to specify the
external format in full detail and to mark which pieces of it are
inessential.  From a QRE, we can infer a regular expression that
describes only the essential information, which we call the internal
format, and we can derive a canonizer that maps between the external
and internal formats.

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}

In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{project} primitive to define this
whitespace-normalizing QRE. \bcp{Why is it called
``project''?} \ksf{select seems like it might be a better name as we
are choosing one representative element from a set of possibilities.}

\begin{lstlisting}
let p_wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

Sometimes there are multiple disjoint representations of the same data.
In such situations, the ARE \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Lovelace, Ada''
and that the \codefont{space\_name} format describes ``Ada Lovelace''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{c\_to\_s} function
is simply to write it from scratch in some ordinary programming language.
However, we can synthesize such functions automatically---here, \codefont{c\_to\_s} is the
\codefont{get} direction of a lens that can be synthesized using the 
\kw{synth} combinator:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
let c_to_s = get l
\end{lstlisting}
\noindent
The first line above synthesizes a lens between \codefont{comma\_name} and
\codefont{space\_name} using the listed example transformation as a guide.  The
second line extracts the \codefont{get} direction transformation from the lens,
which is what we need for \kw{squash}.

Another QRE primitive is the sequential composition combinator,
written ``;''.
For an example of its use, suppose we have already defined a QRE
\codefont{comma_to_space} that accepts names in either space-separated
or name-separated format and chooses the space-separated format as canonical.
Suppose we have also already defined a
QRE \codefont{project_in_space}, which projects away extra white space
between the first and last name in the space-separated name format. 
We can use the sequential composition combinator to combine these two QREs:
\begin{lstlisting}
let name = project_in_space ; comma_to_space
\end{lstlisting}

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp+) $\to$ ",\n")
\end{lstlisting}

The final QRE combinator is the \kw{normalize} combinator. This combinator
allows a programmer to manually define a function $f$ which sends each
string that matches a regular expression $R$ to some canonical
representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence
relation defined by the {\em fibres} of $f$; that is, for all strings
$s$ and $s'$ that match $R$, $s$ is equivalent to $s'$ if and only if
$f(s) = f(s')$.

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p\_wsp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp+, " ", f)}.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. \ksf{Is all this code
correct?  Does it run through the system correctly?  I'm suspicious of
the comma\_to\_space definition which seems to be recurisive.  If it's
correct, we should comment on it.}}
\label{fig:example-qre}
\end{figure}

\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective string lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.
